---
title: "Data Science for Bioinformatics - Week 03"
output: 
  html_document:
    code_folding: show
    toc: yes
editor_options: 
  chunk_output_type: console
---

# Learning goals for this exercise

 * Get more comfortable with Rstudio and R
 * Solve some practice problems in the book (answers are in the back of the book).
 * Perform simple probability calculations in R
 * Make a small function in R
 * Use and understand the sample() function (we use this a LOT)

# Loading library and getting ready.

```{r}

library(tidyverse)

theme_set(theme_classic(base_size=10))

```

# Before you begin - a few tricks 

## Solving probability in R - the elegant way and the brute force approach

"Solve in R" could be: What is the probability of getting TTTT with a fair coin (four tails in a row).

```{r}

prob_tail <- 0.5

(prob_tail^4)

```

Another way of solving/checking is brute force estimating it.

It is ignoring all probability theory...

It is slow...

...





... but?




BUT - it works and it's great in the few cases where you are not able to calculate probabilities (for various reasons).


So another way would be to ask 1 million students to throw a coin 4 times. And then I would count how many of the 1 million students got TTTT.

This will take a long time (AU currently lists 38.000 students).

Luckily we can simulate coin throwing in R

...


So now I will let R throw a coin 4 times, record the outcome - and repeat 1 million times.

In the end I will count how many times I got TTTT

Run the code line by line and notice the output.


```{r}

set.seed(1) # We set the random number generator so we all get identical results.
sample(x = c("H", "T"), size = 4, replace = T) # See what this function does?
sample(x = c("H", "T"), size = 4, replace = T) # See what this function does?
sample(x = c("H", "T"), size = 4, replace = T) # See what this function does?

# Try again - just to see that the set.seed() actually works

set.seed(1)
sample(x = c("H", "T"), size = 4, replace = T)
sample(x = c("H", "T"), size = 4, replace = T)
sample(x = c("H", "T"), size = 4, replace = T)

paste(sample(x = c("H", "T"), size = 4, replace = T), collapse="")

# Make a coin flipping function in R

flip_coin <- function(n) {
  flip <- sample(x = c("H", "T"), size = n, replace = T)
  toss <- paste(flip, collapse="")
  return(toss)
}

set.seed(1)
flip_coin(1)
flip_coin(4)
flip_coin(4)

# Now we will flip a coin 4 times and do it 1 million times

set.seed(0)
y <- rep(NA, 1000000) # We output to a vector - for speed (Remember that it is super fast to write to a vector!!?)

for (i in 1:1000000) {
  y[i] <- flip_coin(4)
  if (i %% 10000 == 0) { # A "progress" bar - print a "." for every 10000 calls
    cat(".")
    flush.console()
  }
}

pd <- tibble(tosses=y) # convert output to a data frame
rm(i,y)

sum(pd$tosses == "HTHT")

```

## Q1.1: What is my brute force estimate of the probability of getting TTTT with a coin using 1 million simulations?


---

## Q1.2: Does it look like that all combinations of four throws have the same probablity?

I.e. does TTTT have the same probability as THTH...

You can use a plot or calculations to answer this.


---

# Book exercises (Analysis of biological data)

## Solve in R the following book exercise as a check 

Do the following practice problems in Chapter 5 to check your knowledge of the simple rules in probability (these will be discussed only briefly in class):

* 1 and 2 (addition and total probability rule) (Brussel sprouts)
* 12 (Five researchers take a random sample)
* 14 (HIV tests)
* 17 (Handwashing)

## NOTE: We ask you to write your answers here in your rmarkdown!!!

When you do that, small calculations can easily be included together with your text/notes.

# Using R to calculate and simulate from known probability distributions

During the course we are going to "meet" a number of theoretical probability distributions, there arise in the context of specific types of data and specific random trials that we imagine are "behind" the data: the 4 distributions we are going to meet in the coming chapters are : 

 * the binomial distribution (chapter 6 and 7 in the ABD book)
 * the Poisson and the chi.square distribution (in chapter 8 )
 * the infamous "normal" or Gaussian distribution (chapters 10, 11 and 12)
 
R is pretty effective at drawing random number and calculating the probability of specific events with these distributions.

So in order to make you comfortable about sampling and the different distributions you we are kindly asking you try it...

## Simulating random numbers from your own distribution / theoretical distributions

Hint:

```{r}

(pd <- tibble(x=1:100))

set.seed(1)
(pd <- tibble(x = sample(c("Palle", "Thomas"), size = 10, replace = TRUE, prob = c(0.8,0.2) )))

set.seed(2)
(pd <- tibble(x = sample(c("Palle", "Thomas"), size = 10, replace = TRUE, prob = c(0.8,0.2) )))

```

## Your task:

Use R functions to make a tibble/data frame with a column full of your random numbers.

Also, make a plot that shows the distribution of your random numbers.

The random numbers should come from:

### Q2.1: A normal distribution with mean 1000 and SD 200

Hint: https://www.google.com/search?q=R+simulate+normal+distribution

Or.. chatGPT or similar? Try it!


---

### Q2.2: A uniform distribution with minimum 10 and maximum 200


---

### Q2.3: A chi square distribution with 1 degrees of freedom

### Q2.4: How big a proportion of your random chi-square values are > 3.84?

(If you had a stats course you may remember this critical value of 3.84)


---

### Q2.5: The empirical distribution of gene length in humans. (The dataset is included)

So we ask you to sample 100 genes (and their lengths) with replacement.

And create a data frame with a column holding the gene lengths as well as visualize the distribution.

```{r}

df <- read_csv(file = "chap04e1HumanGeneLengthsLongestTranscript.csv")

```


---

### Q2.6: Plot the distribution of all the genes. Do you expect the mean or median gene length to be the largest? Why? Check it!


---

# Sampling a NULL distribution from your genes

I have found 56 highly expressed genes with a median length of 3654 - which is longer than the median length of all the genes.

Now, try and make many samples of 56 genes and calculate many median lengths.

## Q3.1: How big a proportion of your samples had a median length >= 3654?

REMEBER THE LECTURE LAST WEEK!! R Speed!!!


---

Congratulations - you have just calculated a one-sided p-value! 

(You just tested if your 56 genes had a significantly longer median length than all genes.)

## Sampling dices

Now we want you to use your knowledge so far to simulate different kind of dices, throw the dice many times and show us that your simulation works!

"Many" is normally at least 10.000...

## Q3.2: A fair 6-sided dice  

Hint:

```{r}

sample(x = 1:4, size = 10, replace = TRUE)

```


---

## Q3.3: An unfair 6-sided dice that never throws 5 (but have equal probability of 1,2,3,4 and 6)


---

## Q3.4: An unfair 6-sided dice that throws 5 with probability 0.1, and the other numbers equal probability. 

In order to check that you do this correctly, you should probably simulate many throws

Hint: ?sample


---

## How often do I get yatzy in a single throw?

## Q3.5: How often do you get yatzy (all 5 dices show the same number of eyes) in one throw? 

Yatzee or yatzy? anyway.... you probably know the game.

So 11111, 222222, 33333, 44444, 55555, 66666 is yatzy.

But how easy is it to get it in a single throw with 5 dices?

You can calculate or simulate


---

## BONUS Q: describe your strategy for simulating yatzy in 3 throws.

This is not super easy. But the rules are:

You have to keep the dices you select to keep.

So 
throw 1: 1,2,4,4,6 - I keep 4,4 and reroll the three other dices trying to get more 4s

throw 2: 4,6,6 - I keep the 4 (now I have three of them) and reroll the remaining two dices

throw 3: 4,1 - I now have 4 4s - but that is not yatzy

Describe your strategy and if you have too much time: code it.

This is not easy. And you are allowed to use any help you can find.



---



