---
title: "Data Science for Bioinformatics - Week 02"
output: 
  html_document:
    code_folding: show
    toc: yes
editor_options: 
  chunk_output_type: console
---

# Learning goals for this exercise

Get a feel for some R functions that we use a lot for data exploring
 
 * dim()
 * nrow()
 * ncol()
 * names()
 * summary()
 * select()
 * filter()
 * group_by()
 * summarise()
 * mutate()

# The mammals dataset

```{r}
library(rstudioapi)

# Getting the path of your current open file
current_path = rstudioapi::getActiveDocumentContext()$path 
setwd(dirname(current_path ))
print( getwd())

library(tidyverse)

mammals <- read_csv(file = "dataset.01.rsbl20150010supp1.csv")

mammals

```

```{r}
mammals = drop_na(mammals)
```

```{r}
library(ggpubr)
```

Note that we used read_csv() to the read the input file.

## Q1.1: Can you use read_delim() to read it? read_tsv()?

```{r}
read_delim(file = "dataset.01.rsbl20150010supp1.csv", delim = ",")

read_tsv(file = "dataset.01.rsbl20150010supp1.csv")
#nope
```

---

## Inspect data

Use dim(), nrow(), ncol(), names(), head() and summary() to inspect the dataset
```{r}
dim(mammals)
nrow(mammals)
ncol(mammals)
names(mammals)
head(mammals)
summary(mammals)
```

## Sidequest: get a quick explanation of all the functions using chatGPT or copilot

they do stuff

---

## Q1.2: How many rows and columns are in the dataset?

17327 rows and 13 columns

---


## Q1.3: Which columns are text?

gene, species, labs, chrMark, chr


---

## Q1.4: Which columns are numbers?
 dN, dNdS, RPKM, Tau, GC3

---

# Gene expression in the mammals dataset

## Q2.1: Vizualize the distribution of gene expression (RPKM) in the dataset

Hint: Make a new code chunk and make a histogram that shows the distribution of RPKM when you ignore species information.

Also, what is the minimum and maximum RPKM value in the dataset?

```{r}
ggplot(data = mammals, aes(x = RPKM)) + geom_histogram() + xlim(0, 1000)
```
```{r}
mammals[mammals$RPKM > 1000, ]
```

---

## Q2.2: Make a new variable called "expression" that is the log10(RPKM+1)

Hint: use the mutate function


```{r}
mammals %>%
	mutate(expression = log10(RPKM + 1)) %>%
	ggplot() +
	geom_histogram(mapping = aes(x = expression))
```

---

## Q2.3: Visualize the overall distribution of the expression but for each species separatedly?

Hint: Make histograms and use facet_wrap() to make subplots.

```{r}
mammals %>%
	mutate(expression = log10(RPKM + 1)) %>%
	ggplot() +
	geom_histogram(mapping = aes(x = expression)) +
	facet_wrap(~Species)
```



---

## Q2.4: Visualize the overall distribution of the expression in humans and mouse.

You probably need a filter() - and then you can choose how you will present it.

Also: we can compare things using different operators.

You can find them with

?Comparison

And the special %in% is described here:

?match

.... I use %in% a lot!


```{r}

"Hello" %in% c("Hello", "world")
"Hello" %in% c("Goodbye", "cruel", "world")

```

```{r}
mammals %>%
	filter(Species %in% c("Human", "Mouse")) %>%
	mutate(expression = log10(RPKM + 1)) -> mammalsfilter
```

```{r}
mammalsfilter %>%
	ggplot() +
	geom_histogram(mapping = aes(x = expression))
```

---

## Q2.5: What is the mean, median and standard deviation of the expression for each Species?

Hint: mutate, group_by and summarise

```{r}
mammals  %>% group_by(Species) %>% summarise(mean = mean(RPKM), median = median(RPKM), sd = sd(RPKM))
```
---



# Open question

Some open questions that you can answer (pick a single one - or come up with your own question!)

 * Are dNdS values the same for all species?

 * Are high dNdS genes sitting on autosomes or X/Y chromosomes?

 * Are genes expression the same across species?

If you can't really solve these - then think about what you would like to do: i.e. describe in words what you would like your code to do.

Again, you are welcome to use chatGPT and/or google to help you.




